Comparing Machine Learning algorithms for classifying cognitive states from fMRI data

### 2012-11-11 article/article.tex
preprocessing

\subsection{Preprocessing}
\label{ssec:preprocessing}
The dataset contains activation values for each voxels with respect to a baseline activation value for a subject, measured before performing the task. To make optimal use of the information value of each voxel, we have created normalized copies of the dataset to use as input for the algorithms. The effect of this preprocessing step depends on assumptions made about the form of our data, which are not available. Instead, we had to verify the effect by looking at the resulting classification results (cf.~Table~\ref{tab:results}). As a first step, we applied row normalization by setting the mean activation value to zero for each trial. The assumption was that the relative activation has more predictive power than the actual value. We then continued by transforming these values to z-scores (by setting the standard-deviation to 1) but this turned out to negatively impact accuracy, indicating that the transformation lost information. For completeness, we also performed normalization of higher-order properties (such as skew and kurtosis) but found a significant further decrease in accuracy. We therefore did not further researched performance on these sets. One other approach we tried was the application of column normalization, setting the mean value across trials to zero.

@@

\emph{ROI mapping.} ROI mapping is a form of feature selection, wherein only a subset of the N-dimensional input space is used. As described in the introduction, ROIs are anatomically defined brain areas classified as relevant by an expert in the field. We used 7 of the 25 defined ROIs, labeled  'CALC' 'LIPL' 'LT' 'LTRIA' 'LOPER' 'LIPS' and  'LDLPFC'. These ROIs have previously been found to be most relevant in discriminating between the cognitive states of looking at sentences or pictures \cite{wa:2003betw}. To further reduce the number of features, average voxel values within these regions are taken as new input features. As other higher-order features such as the standard deviation, kurtosis and skewness within ROIs did not increase predictability, these were not included in the subsequent analysis. The disadvantage of using ROIs is that they require \textit{a priori} knowledge. When this knowledge is inaccurate, classification accuracy may not be optimal.

@@

where $d$ denotes the dimensionality of $x$ and $\Sigma$ denotes the covariance matrix. The parameters are estimated per scan and the decision method is similar to GNB. The increase in parameters to be estimated compared to GNB (the correlations) may decrease generality, while also decreasing inductive bias.

>>>>

\subsection{Preprocessing}
\label{ssec:preprocessing}
The dataset contains activation values for each voxels with respect to a baseline activation value for a subject, measured before performing the task. To make optimal use of the information value of each voxel, we have created normalized copies of the dataset to use as input for the algorithms. The effect of this preprocessing step depends on assumptions made about the form of our data, which are not available. Instead, we had to verify the effect by looking at the resulting classification results (cf.~Table~\ref{tab:results}).

As a first step, we applied row normalization by setting the mean activation value to zero for each trial. The assumption was that the relative activation has more predictive power than the actual value. We then continued by transforming these values to z-scores by setting the standard-deviation to 1. However, this turned out to negatively impact accuracy, indicating that the transformation lost information. For completeness, we also performed normalization of higher-order properties, such as skew and kurtosis, but found a significant further decrease in accuracy. We therefore did not further researched performance on these sets. One other approach we tried was the application of column normalization, setting the mean value across trials to zero.

@@

\textbf{\emph{ROI mapping.}} ROI mapping is a form of feature selection, wherein only a subset of the N-dimensional input space is used. As described in the introduction, ROIs are anatomically defined brain areas classified as relevant by an expert in the field. We used 7 of the 25 defined ROIs, labeled  'CALC' 'LIPL' 'LT' 'LTRIA' 'LOPER' 'LIPS' and  'LDLPFC'. These ROIs have previously been found to be most relevant in discriminating between the cognitive states of looking at sentences or pictures \cite{wa:2003betw}. To further reduce the number of features, average voxel values within these regions are taken as new input features. As other higher-order features such as the standard deviation, kurtosis and skewness within ROIs did not increase predictability, these were not included in the subsequent analysis. The disadvantage of using ROIs is that they require \textit{a priori} knowledge. When this knowledge is inaccurate, classification accuracy may not be optimal.

@@

where $d$ denotes the dimensionality of $x$ and $\Sigma$ denotes the covariance matrix. The parameters are estimated per scan and the decision method is similar to GNB. The increase in parameters to be estimated compared to GNB (the correlations) may decrease generalization, while also decreasing inductive bias.
